---
title: "Full Race Cross Validation"
author: "Cole Cappello"
date: "09-24-25
output: pdf_document
---

```{r}

knitr::opts_chunk$set(message = FALSE)

set.seed(126)

knitr::opts_chunk$set(echo = TRUE)
library(mosaic)
library(tidyverse)
library(effects)
library(ggResidpanel)
library(catstats2)
library(Matrix)
library(KFAS)
library(rstan)
library(loo)

options(mc.cores = parallel::detectCores())

```

# Simulation

We'll create a fuel.kg vector which starts at 110 (this is the max amount of fuel allowed by regulation) and decreases linearly to 0 at the end of the race. We are going to add a covariate to the observation equation for fuel and estimate its effect gamma. In order to estimate gamma correctly, it will be critical to set a proper prior for the initial latent $\alpha_1 = \alpha.reset[compound_{t=1}]$. Ultimately, it needs to be roughly 1-1.5 seconds less than the initial observation $y_1$, but in the stan code we'll set the prior to be centered on the true value of 95.

\begin{align}
	y_t &= \alpha_t + \gamma*fuel.kg_t +\epsilon_t \\
	\alpha_{t+1} &= 
	\begin{cases}
		\alpha_t + \nu_t + \omega_t & pit_t = 0 \\ 
		\alpha.reset[compound_t] & pit = 1
	\end{cases} \\	
	\nu_{t+1} &= 
	\begin{cases}
		\nu_t + \beta[compound_t] & pit_t = 0 \\
		\nu.reset & pit = 1
	\end{cases}	
\end{align}

where \( \epsilon_t \sim N(0,.2) \), \( \beta[Hard] = .002 \), $\beta[Medium] = .01$, $\beta[Soft] = .02$, $\nu_1 = \nu.reset = 0$, $\omega_t \sim N(0,.15)$, and $\alpha_1 = \alpha.reset[compound_{t=1}]$.  We'll let $\alpha.reset$ be 95, 94.5, and 94 for hard, medium, and soft tires resepectively. In this simulation, we will start on hard tires, pit on lap 30 for soft tires, then pit on lap 40 for medium tires. 

```{r}

# Number of states/laps
N <- 70

# Tire Compound index 1 - Hard, 2 - Medium, 3 - Soft
compound <- c(rep(1,times = 30), rep(3, times = 20),rep(2, times = 20))

# Fuel Mass in kg
fuel.kg <- seq(from = 110, to = 0, length.out = N)

# Gamma 
gamma <- .01


# Pit Index indicates with a 1 if a new set of tires is put on the next lap
pit <- rep(0, N)
pit[30] <- 1
pit[50] <- 1

# Beta is now a vector with the index representing each compound
beta <- rep(0,3)
beta[1] <- .002
beta[2] <- .01
beta[3] <- .02

# Reset vector stores the true pace of a new tire for each compound
reset.alpha <- rep(0,3)
reset.alpha[1] <- 95
reset.alpha[2] <- 94.5
reset.alpha[3] <- 94

# Reset value for nu
reset.nu <- 0
    
# Initialize nu vector
nu <- rep(NA, N)
nu[1] <- .01

# Initialize alpha vector
alpha <- rep(NA, N)
alpha[1] <- reset.alpha[1]

# Initialize process errors
omega <- rnorm(n = N, mean = 0, sd = .1)

# Store true latent states
for(i in 2:(length(alpha))) {
    if(pit[i-1] == 1) {
        alpha[i] <- reset.alpha[compound[i]] + omega[i]
        nu[i] <- reset.nu
    } else {
    alpha[i] <- alpha[i-1] + nu[i-1] + omega[i]
    nu[i] <- nu[i-1] + beta[compound[i-1]]
    }
}    

# Vector of observation errors
epsilon <- rnorm(n = N, mean = 0, sd = .2)

# Simulated observations
y <- alpha + fuel.kg*gamma + epsilon

plot(x = 1:N, y = y)

```

## Cross Validation Code for Stan Models 

Here things are going to get messy.  Since there are multiple stints, we have to do cross validation on the data points at the end of each stint.

```{r}


stint_CV <- function(model_string, data_list) {

    # Store the full data/time series
    y <- data_list[["y"]]
    # Store the full compound vector
    compound <- data_list[["Compound"]]
    # Store the full pit vector
    pit <- data_list[["Pit"]]
    # Store the Fuel vector
    fuel <- data_list[["fuel_mass"]]

   
    # Get the index of laps that end a stint
    end_stint <- which(data_list[["Pit"]] == 1)
    # Add the last lap because that also marks the end of a stint
    end_stint <- c(end_stint,length(y))

    # Number of Stints
    num_stints <- length(end_stint)
    
    # Will get RMSPE for each stint so need an empty vector
    RMSPE_vec <- rep(NA, num_stints)

    # Will get plots for each stint
    plots <- vector("list", length = num_stints)
    
    for(j in 1:num_stints) {

        # Do CV on last quarter of each stint        
        # Number of folds
        if(j == 1) {
            K <- round(end_stint[j]/5)
        } else {
            K <- round((end_stint[j] - end_stint[j-1])/5)
        }    

        stan.models <- vector("list", length = K)
        posteriors <- vector("list", length = K)

        # Vector to Store one-step ahead predictions
        ypred_hat <- rep(NA, K)

        for(i in 1:K) {

            # Update data vector to allow for next step prediction
            data_list[["TT"]] <- length(y[1:(i+end_stint[j]-K-1)])
            data_list[["y"]] <- y[1:(i+end_stint[j]-K-1)]
            data_list[["Compound"]] <- compound[1:(i+end_stint[j]-K-1)]
            c_map <- sort(unique(compound[1:(i+end_stint[j]-K-1)]))
            data_list[["compound_map"]] <- array(c_map, dim = length(c_map))
            data_list[["C_used"]] <- length(c_map)
            data_list[["Pit"]] <- pit[1:(i+end_stint[j]-K-1)]
            data_list[["fuel_mass"]] <- fuel[1:(i+end_stint[j]-K-1)]
            
            # Fit model on subset of data then get one step ahead prediction
            stan.models[[i]] <- stan(file = model_string,
                                     data = data_list,
                                     chains = 3, iter = 30000,
                                     control = list(adapt_delta = .99, max_treedepth = 12))
            posteriors[[i]] <- extract(stan.models[[i]])
            ypred_hat[i]  <- mean(posteriors[[i]]$y_pred)
            
        }    

        # Plots
        df <- tibble(laps = 1:end_stint[j],
                     y = y[1:end_stint[j]])

        pred_df <- tibble(time = (end_stint[j] - K + 1):end_stint[j],
                          pred = ypred_hat)

        plots[[j]] <- ggplot() +
            geom_point(data = df,
                       aes(x = laps, y = y, color = "Observations")) +
            geom_point(data = pred_df,
                       aes(x = time, y = pred, color = "Predictions")) +
            scale_color_manual(values = c("Observations" = "red",
                                          "Predictions" = "blue")) +
            labs(title = paste("Scatter Plot of Lap vs Lap Times with Predictions for Stint", j),
                 x = "Lap",
                 y = "Laptime in Seconds",
                 color = "Legend")
        
        # Calculate Summary Statistic
        MSPE <- mean((y[(end_stint[j]-K+1):end_stint[j]] - ypred_hat)^2)
        RMSPE_vec[j] <- sqrt(MSPE)

    }
    return(list(RMSPE_vec,plots))
}
# list(TT = length(y[1:(i+N-K-1)]), y = y[1:(i+N-K-1)], z0 = 96, v0 = .01, sdo0 = .3)

```

## Cross Validation Code for Arima Baseline

```{r}

arima_cv <- function(y, pit) {

    # Get the index of laps that end a stint
    end_stint <- which(pit == 1)
    # Add the last lap because that also marks the end of a stint
    end_stint <- c(end_stint,length(y))

    # Number of Stints
    num_stints <- length(end_stint)
    
    # Will get RMSPE for each stint so need an empty vector
    RMSPE_vec <- rep(NA, num_stints)

    # Will get plots for each stint
    plots <- vector("list", length = num_stints)

    for(i in 1:num_stints) {

        # Number of folds
        if(i == 1) {
            K <- round(end_stint[i]/5)
        } else {
            K <- round((end_stint[i] - end_stint[i-1])/5)
        }    

        arma.models <- vector("list", length = K)

        # Vector to Store one-step ahead predictions
        ypred_hat <- rep(NA, K)

        for(j in 1:K) {

            arma.models[[j]] <- arima(y[1:(j+end_stint[i]-K-1)], order = c(3,2,3))
            ypred_hat[j] <- as.numeric(predict(arma.models[[j]],n.ahead = 1)$pred)
            
        }
                # Plots
        df <- tibble(laps = 1:end_stint[i],
                     y = y[1:end_stint[i]])

        pred_df <- tibble(time = (end_stint[i] - K + 1):end_stint[i],
                          pred = ypred_hat)

        plots[[i]] <- ggplot() +
            geom_point(data = df,
                       aes(x = laps, y = y, color = "Observations")) +
            geom_point(data = pred_df,
                       aes(x = time, y = pred, color = "Predictions")) +
            scale_color_manual(values = c("Observations" = "red",
                                          "Predictions" = "blue")) +
            labs(title = paste("Scatter Plot of Lap vs Lap Times with Predictions for Stint", i),
                 x = "Lap",
                 y = "Laptime in Seconds",
                 color = "Legend")
        
        # Calculate Summary Statistic
        MSPE <- mean((y[(end_stint[i]-K+1):end_stint[i]] - ypred_hat)^2)
        RMSPE_vec[i] <- sqrt(MSPE)
    }

    return(list(RMSPE_vec,plots))
    
}



```



# CV for Various Models

```{r}

data_full <- list(TT = length(y), y = y, C = 3, Compound = compound, C_used = 3, compound_map = c(1,2,3), Pit = pit, z_reset0 = c(95,94.5,94), v_reset0 = 0, sdo0 = .3, fuel_mass = fuel.kg)


RMSPE.vec <- stint_CV("full_race_1driver_and_fuel_real.stan", data_full)

RMSPE.vec[[2]][[1]]

RMSPE.vec[[2]][[2]]

RMSPE.vec[[2]][[3]]


# CV RMSPE
RMSPE.vec[[1]]


attempt1 <- arima_cv(y,pit)

```


