---
title: "Cross Validation on a Real Race"
author: "Cole Cappello"
date: "09-29-25"
output: pdf_document
---

```{r, message=FALSE}

# Setup
library(tidyverse)
library(Matrix)
library(rstan)

options(mc.cores = parallel::detectCores())

```

# The Hugnarian Grand Prix 2025

I'm going to load lap time data from the Hungarian Grand Prix for various drivers.  We'll do cross validation on Charles Leclerc's race because - while I admit I am slightly partial - he does seem to have a race that conforms relatively well to our model. Here is a visual of Leclerc's lap times:

```{r, message=FALSE}

hungary_laps <- read_csv("Hungary_laps.csv")

hungary_laps <- hungary_laps %>%
    mutate(Stint = as.factor(Stint),
           Compound = as.factor(Compound))

filter(hungary_laps, Driver == "LEC", is.na(PitOutTime) & is.na(PitInTime),LapTime < 100) %>%
    ggplot(mapping = aes(x = LapNumber, y = LapTime, colour = Stint)) +
    geom_point() +
    geom_smooth() +
    labs(title = "Tire Degration -- Charles Leclerc (Ferrari)",
         x = "Lap Number",
         y = "Lap Time (Seconds)")

```


# Cross Validation Code

In this section I will include the code used for cross validation. Skip to the next section if interested in results.

## Stan CV code

```{r, message=FALSE, results='hide', warning=FALSE}

stint_CV <- function(model_string, data_list) {

    # Store the full data/time series
    y <- data_list[["y"]]
    # Store the full compound vector
    compound <- data_list[["Compound"]]
    # Store the full pit vector
    pit <- data_list[["Pit"]]
    # Store the Fuel vector
    fuel <- data_list[["fuel_mass"]]

   
    # Get the index of laps that end a stint
    end_stint <- which(data_list[["Pit"]] == 1)
    # Add the last lap because that also marks the end of a stint
    end_stint <- c(end_stint,length(y))

    # Number of Stints
    num_stints <- length(end_stint)
    
    # Will get RMSPE for each stint so need an empty vector
    RMSPE_vec <- rep(NA, num_stints)

    # Will get plots for each stint
    plots <- vector("list", length = num_stints)
    
    for(j in 1:num_stints) {

        # Do CV on last quarter of each stint        
        # Number of folds
        if(j == 1) {
            K <- round(end_stint[j]/4)
        } else {
            K <- round((end_stint[j] - end_stint[j-1])/4)
        }    

        stan.models <- vector("list", length = K)
        posteriors <- vector("list", length = K)

        # Vector to Store one-step ahead predictions
        ypred_hat <- rep(NA, K)

        for(i in 1:K) {

            # Update data vector to allow for next step prediction
            data_list[["TT"]] <- length(y[1:(i+end_stint[j]-K-1)])
            data_list[["y"]] <- y[1:(i+end_stint[j]-K-1)]
            data_list[["Compound"]] <- compound[1:(i+end_stint[j]-K-1)]
            c_map <- sort(unique(compound[1:(i+end_stint[j]-K-1)]))
            data_list[["compound_map"]] <- array(c_map, dim = length(c_map))
            data_list[["C_used"]] <- length(c_map)
            data_list[["Pit"]] <- pit[1:(i+end_stint[j]-K-1)]
            data_list[["fuel_mass"]] <- fuel[1:(i+end_stint[j]-K-1)]
            
            # Fit model on subset of data then get one step ahead prediction
            stan.models[[i]] <- stan(file = model_string,
                                     data = data_list,
                                     chains = 3, iter = 30000,
                                     control = list(adapt_delta = .99, max_treedepth = 12))
            posteriors[[i]] <- extract(stan.models[[i]])
            ypred_hat[i]  <- mean(posteriors[[i]]$y_pred)
            
        }    

        # Plots
        df <- tibble(laps = 1:end_stint[j],
                     y = y[1:end_stint[j]])

        pred_df <- tibble(time = (end_stint[j] - K + 1):end_stint[j],
                          pred = ypred_hat)

        plots[[j]] <- ggplot() +
            geom_point(data = df,
                       aes(x = laps, y = y, color = "Observations")) +
            geom_point(data = pred_df,
                       aes(x = time, y = pred, color = "Predictions")) +
            scale_color_manual(values = c("Observations" = "red",
                                          "Predictions" = "blue")) +
            labs(title = paste("Scatter Plot of Lap vs Lap Times with Predictions for Stint", j),
                 x = "Lap",
                 y = "Laptime in Seconds",
                 color = "Legend")
        
        # Calculate Summary Statistic
        MSPE <- mean((y[(end_stint[j]-K+1):end_stint[j]] - ypred_hat)^2)
        RMSPE_vec[j] <- sqrt(MSPE)

    }
    return(list(RMSPE_vec,plots))
}

```


## Arima Baseline CV code

```{r, message=FALSE, results='hide'}

arima_cv <- function(y, pit) {

    # Get the index of laps that end a stint
    end_stint <- which(pit == 1)
    # Add the last lap because that also marks the end of a stint
    end_stint <- c(end_stint,length(y))

    # Number of Stints
    num_stints <- length(end_stint)
    
    # Will get RMSPE for each stint so need an empty vector
    RMSPE_vec <- rep(NA, num_stints)

    # Will get plots for each stint
    plots <- vector("list", length = num_stints)

    for(i in 1:num_stints) {

        # Number of folds
        if(i == 1) {
            K <- round(end_stint[i]/4)
        } else {
            K <- round((end_stint[i] - end_stint[i-1])/4)
        }    

        arma.models <- vector("list", length = K)

        # Vector to Store one-step ahead predictions
        ypred_hat <- rep(NA, K)

        for(j in 1:K) {

            arma.models[[j]] <- arima(y[1:(j+end_stint[i]-K-1)], order = c(2,1,2))
            ypred_hat[j] <- as.numeric(predict(arma.models[[j]],n.ahead = 1)$pred)
            
        }
                # Plots
        df <- tibble(laps = 1:end_stint[i],
                     y = y[1:end_stint[i]])

        pred_df <- tibble(time = (end_stint[i] - K + 1):end_stint[i],
                          pred = ypred_hat)

        plots[[i]] <- ggplot() +
            geom_point(data = df,
                       aes(x = laps, y = y, color = "Observations")) +
            geom_point(data = pred_df,
                       aes(x = time, y = pred, color = "Predictions")) +
            scale_color_manual(values = c("Observations" = "red",
                                          "Predictions" = "blue")) +
            labs(title = paste("Scatter Plot of Lap vs Lap Times with Predictions for Stint", i),
                 x = "Lap",
                 y = "Laptime in Seconds",
                 color = "Legend")
        
        # Calculate Summary Statistic
        MSPE <- mean((y[(end_stint[i]-K+1):end_stint[i]] - ypred_hat)^2)
        RMSPE_vec[i] <- sqrt(MSPE)
    }

    return(list(RMSPE_vec,plots))
    
}

```

# Analysis

## Get the data ready

```{r}

leclerc_df <- hungary_laps %>%
    filter(Driver == "LEC", is.na(PitOutTime) & is.na(PitInTime), LapTime < 100) %>%
    mutate(Compound_Code = as.integer(factor(Compound,levels = c("HARD","MEDIUM","SOFT"))))

leclerc_laps <- leclerc_df$LapTime

leclerc_compound <- leclerc_df$Compound_Code

leclerc_pit <- rep(0,length(leclerc_compound))

for(i in 2:length(leclerc_df$Stint)) {
    if(leclerc_df$Stint[i] != leclerc_df$Stint[i-1]) {
        leclerc_pit[i-1] <- 1
    }    
}                  

fuel.kg <- seq(110,1,length.out = length(leclerc_laps))

leclerc_map <- sort(unique(leclerc_compound))
leclerc_used <- length(leclerc_map)

```

## Cross Validation Results

```{r, warning=FALSE}

# Arima cv
arima_results <- arima_cv(leclerc_laps, leclerc_pit)

# Stan SSM CV
leclerc_data <- list(TT = length(leclerc_laps), y = leclerc_laps, C = 3, Compound = leclerc_compound, C_used = leclerc_used, compound_map = leclerc_map, Pit = leclerc_pit, z_reset0 = c(80.5,79,79), v_reset0 = 0, sdo0 = .3, fuel_mass = fuel.kg)

stan_results <- stint_CV("full_race_1driver_and_fuel_real.stan", leclerc_data)

```


```{r}

# Stint 1
arima_results[[1]][[1]]
stan_results[[1]][[1]]

# Stint 2
arima_results[[1]][[2]]
stan_results[[1]][[2]]

#Stint 3
arima_results[[1]][[3]]
stan_results[[1]][[3]]

```



# Appendix

Plots of lap time vs lap with observations and predictions from both the arima model and our full SSM.

## Arima Predictions

```{r}

arima_results[[2]][[1]]

arima_results[[2]][[2]]

arima_results[[2]][[3]]

```

## SSM Predictions

```{r}

stan_results[[2]][[1]]

stan_results[[2]][[2]]

stan_results[[2]][[3]]

```



