---
title: "Cross Validation on a Real Race"
author: "Cole Cappello"
date: "09-29-25"
output: pdf_document
---

```{r, message=FALSE}

# Setup
library(tidyverse)
library(Matrix)
library(rstan)

options(mc.cores = parallel::detectCores())

```

# The Austrian Grand Prix 2025

I'm going to load lap time data from the Austrian Grand Prix for various drivers.  We'll do cross validation on Lewis Hamilton's race as he does seem to have a race that conforms relatively well to our model. Here is a visual of Hamilton's lap times:

```{r, message=FALSE}

austria_laps <- read_csv("Austria_laps.csv")

austria_laps <- austria_laps %>%
    mutate(Stint = as.factor(Stint),
           Compound = as.factor(Compound))

filter(austria_laps, Driver == "HAM", is.na(PitOutTime) & is.na(PitInTime),LapTime < 100) %>%
    ggplot(mapping = aes(x = LapNumber, y = LapTime, colour = Stint)) +
    geom_point() +
    geom_smooth() +
    labs(title = "Tire Degration -- Lewis Hamilton (Ferrari)",
         x = "Lap Number",
         y = "Lap Time (Seconds)")

```


# Cross Validation Code

In this section I will include the code used for cross validation. Skip to the next section if interested in results.

## Stan CV code

```{r, message=FALSE, results='hide', warning=FALSE}

stint_CV <- function(model_string, data_list) {

    # Store the full data/time series
    y <- data_list[["y"]]
    # Store the full compound vector
    compound <- data_list[["Compound"]]
    # Store the full pit vector
    pit <- data_list[["Pit"]]
    # Store the Fuel vector
    fuel <- data_list[["fuel_mass"]]

   
    # Get the index of laps that end a stint
    end_stint <- which(data_list[["Pit"]] == 1)
    # Add the last lap because that also marks the end of a stint
    end_stint <- c(end_stint,length(y))

    # Number of Stints
    num_stints <- length(end_stint)
    
    # Will get RMSPE for each stint so need an empty vector
    RMSPE_vec <- rep(NA, num_stints)

    # Will get plots for each stint
    plots <- vector("list", length = num_stints)
    
    for(j in 1:num_stints) {

        # Do CV on last quarter of each stint        
        # Number of folds
        if(j == 1) {
            K <- round(end_stint[j]/4)
        } else {
            K <- round((end_stint[j] - end_stint[j-1])/4)
        }    

        stan.models <- vector("list", length = K)
        posteriors <- vector("list", length = K)

        # Vector to Store one-step ahead predictions
        ypred_hat <- rep(NA, K)

        for(i in 1:K) {

            # Update data vector to allow for next step prediction
            data_list[["TT"]] <- length(y[1:(i+end_stint[j]-K-1)])
            data_list[["y"]] <- y[1:(i+end_stint[j]-K-1)]
            data_list[["Compound"]] <- compound[1:(i+end_stint[j]-K-1)]
            c_map <- sort(unique(compound[1:(i+end_stint[j]-K-1)]))
            data_list[["compound_map"]] <- array(c_map, dim = length(c_map))
            data_list[["C_used"]] <- length(c_map)
            data_list[["Pit"]] <- pit[1:(i+end_stint[j]-K-1)]
            data_list[["fuel_mass"]] <- fuel[1:(i+end_stint[j]-K-1)]
            
            # Fit model on subset of data then get one step ahead prediction
            stan.models[[i]] <- stan(file = model_string,
                                     data = data_list,
                                     chains = 3, iter = 30000,
                                     control = list(adapt_delta = .99, max_treedepth = 12))
            posteriors[[i]] <- extract(stan.models[[i]])
            ypred_hat[i]  <- mean(posteriors[[i]]$y_pred)
            
        }    

        # Plots
        df <- tibble(laps = 1:end_stint[j],
                     y = y[1:end_stint[j]])

        pred_df <- tibble(time = (end_stint[j] - K + 1):end_stint[j],
                          pred = ypred_hat)

        plots[[j]] <- ggplot() +
            geom_point(data = df,
                       aes(x = laps, y = y, color = "Observations")) +
            geom_point(data = pred_df,
                       aes(x = time, y = pred, color = "Predictions")) +
            scale_color_manual(values = c("Observations" = "red",
                                          "Predictions" = "blue")) +
            labs(title = paste("Scatter Plot of Lap vs Lap Times with Predictions for Stint", j),
                 x = "Lap",
                 y = "Laptime in Seconds",
                 color = "Legend")
        
        # Calculate Summary Statistic
        MSPE <- mean((y[(end_stint[j]-K+1):end_stint[j]] - ypred_hat)^2)
        RMSPE_vec[j] <- sqrt(MSPE)

    }
    return(list(RMSPE_vec,plots))
}

```


## Arima Baseline CV code

```{r, message=FALSE, results='hide'}

arima_cv <- function(y, pit) {

    # Get the index of laps that end a stint
    end_stint <- which(pit == 1)
    # Add the last lap because that also marks the end of a stint
    end_stint <- c(end_stint,length(y))

    # Number of Stints
    num_stints <- length(end_stint)
    
    # Will get RMSPE for each stint so need an empty vector
    RMSPE_vec <- rep(NA, num_stints)

    # Will get plots for each stint
    plots <- vector("list", length = num_stints)

    for(i in 1:num_stints) {

        # Number of folds
        if(i == 1) {
            K <- round(end_stint[i]/4)
        } else {
            K <- round((end_stint[i] - end_stint[i-1])/4)
        }    

        arma.models <- vector("list", length = K)

        # Vector to Store one-step ahead predictions
        ypred_hat <- rep(NA, K)

        for(j in 1:K) {

            arma.models[[j]] <- arima(y[1:(j+end_stint[i]-K-1)], order = c(2,1,2))
            ypred_hat[j] <- as.numeric(predict(arma.models[[j]],n.ahead = 1)$pred)
            
        }
                # Plots
        df <- tibble(laps = 1:end_stint[i],
                     y = y[1:end_stint[i]])

        pred_df <- tibble(time = (end_stint[i] - K + 1):end_stint[i],
                          pred = ypred_hat)

        plots[[i]] <- ggplot() +
            geom_point(data = df,
                       aes(x = laps, y = y, color = "Observations")) +
            geom_point(data = pred_df,
                       aes(x = time, y = pred, color = "Predictions")) +
            scale_color_manual(values = c("Observations" = "red",
                                          "Predictions" = "blue")) +
            labs(title = paste("Scatter Plot of Lap vs Lap Times with Predictions for Stint", i),
                 x = "Lap",
                 y = "Laptime in Seconds",
                 color = "Legend")
        
        # Calculate Summary Statistic
        MSPE <- mean((y[(end_stint[i]-K+1):end_stint[i]] - ypred_hat)^2)
        RMSPE_vec[i] <- sqrt(MSPE)
    }

    return(list(RMSPE_vec,plots))
    
}

```

# Analysis

## Get the data ready

```{r}

leclerc_df <- austria_laps %>%
    filter(Driver == "LEC", is.na(PitOutTime) & is.na(PitInTime), LapTime < 100) %>%
    mutate(Compound_Code = as.integer(factor(Compound,levels = c("HARD","MEDIUM","SOFT"))))

leclerc_laps <- leclerc_df$LapTime

leclerc_compound <- leclerc_df$Compound_Code

leclerc_pit <- rep(0,length(leclerc_compound))

for(i in 2:length(leclerc_df$Stint)) {
    if(leclerc_df$Stint[i] != leclerc_df$Stint[i-1]) {
        leclerc_pit[i-1] <- 1
    }    
}                  

fuel.kg <- seq(110,1,length.out = length(leclerc_laps))

leclerc_map <- sort(unique(leclerc_compound))
leclerc_used <- length(leclerc_map)

```

## Cross Validation Results

```{r, warning=FALSE}

# Arima cv
arima_results <- arima_cv(leclerc_laps, leclerc_pit)

# Stan SSM CV
leclerc_data <- list(TT = length(leclerc_laps), y = leclerc_laps, C = 3, Compound = leclerc_compound, C_used = leclerc_used, compound_map = leclerc_map, Pit = leclerc_pit, z_reset0 = c(80.5,79,79), v_reset0 = 0, sdo0 = .3, fuel_mass = fuel.kg)

stan_results <- stint_CV("full_race_1driver_and_fuel_real.stan", leclerc_data)

```


```{r}

# Stint 1
arima_results[[1]][[1]]
stan_results[[1]][[1]]

# Stint 2
arima_results[[1]][[2]]
stan_results[[1]][[2]]

#Stint 3
arima_results[[1]][[3]]
stan_results[[1]][[3]]

```



# Appendix

Plots of lap time vs lap with observations and predictions from both the arima model and our full SSM.

## Arima Predictions

```{r}

arima_results[[2]][[1]]

arima_results[[2]][[2]]

arima_results[[2]][[3]]

```

## SSM Predictions

```{r}

stan_results[[2]][[1]]

stan_results[[2]][[2]]

stan_results[[2]][[3]]

```

# Hamilton Code

## Tests

```{r}

hamilton_df <- austria_laps %>%
    filter(Driver == "HAM", is.na(PitOutTime) & is.na(PitInTime), LapTime < 100) %>%
    mutate(Compound_Code = as.integer(factor(Compound,levels = c("HARD","MEDIUM","SOFT"))))

hamilton_laps <- hamilton_df$LapTime

hamilton_compound <- hamilton_df$Compound_Code

hamilton_pit <- rep(0,length(hamilton_compound))

for(i in 2:length(hamilton_df$Stint)) {
    if(hamilton_df$Stint[i] != hamilton_df$Stint[i-1]) {
        hamilton_pit[i-1] <- 1
    }    
}                  

fuel.kg <- seq(110,1,length.out = length(hamilton_laps))

hamilton_map <- sort(unique(hamilton_compound))
hamilton_used <- length(hamilton_map)

# Data for stan code
hamilton_data <- list(TT = length(hamilton_laps), y = hamilton_laps, C = 3, Compound = hamilton_compound, C_used = hamilton_used, compound_map = hamilton_map, Pit = hamilton_pit, z_reset0 = 69, v_reset0 = 0, sdo0 = .3, fuel_mass = fuel.kg)

test_model <- stan(file = "full_race_1driver_and_fuel_base.stan",
                        data = hamilton_data,
                        chains = 3, iter = 30000,
                        control = list(adapt_delta = .99, max_treedepth = 12))

posterior_test <- extract(test_model)

alphas <- colMeans(posterior_test[["z"]])

plotting <- tibble(Lap = 1:63,
                   alphas = alphas,
                   observations = hamilton_laps)

ggplot(plotting, mapping = aes(x = Lap)) + 
  geom_point(mapping = aes(y =observations, col = "Observations"))+
  geom_point(mapping = aes(y = alphas, col = "Latent States"))



##########################################################################################

# Data for stan code
hamilton_data <- list(TT = length(hamilton_laps), y = hamilton_laps, C = 3, Compound = hamilton_compound, C_used = hamilton_used, compound_map = hamilton_map, Pit = hamilton_pit, z_reset0 = c(69.5,69,69.5), v_reset0 = 0, sdo0 = .3, fuel_mass = fuel.kg)

test_model <- stan(file = "full_race_1driver_and_fuel_Extension1.stan",
                        data = hamilton_data,
                        chains = 3, iter = 30000,
                        control = list(adapt_delta = .99, max_treedepth = 12))

posterior_test <- extract(test_model)

alphas <- colMeans(posterior_test[["z"]])

plotting <- tibble(Lap = 1:63,
                   alphas = alphas,
                   observations = hamilton_laps)

ggplot(plotting, mapping = aes(x = Lap)) + 
  geom_point(mapping = aes(y =observations, col = "Observations"))+
  geom_point(mapping = aes(y = alphas, col = "Latent States"))


##########################################################################################

hamilton_data <- list(TT = length(hamilton_laps), y = hamilton_laps, C = 3, Compound = hamilton_compound, C_used = hamilton_used, compound_map = hamilton_map, Pit = hamilton_pit, z_reset0 = c(69,68.5,68), v_reset0 = 0, sdo0 = .3, fuel_mass = fuel.kg)

test_model <- stan(file = "full_race_1driver_and_fuel_Extension2.stan",
                        data = hamilton_data,
                        chains = 3, iter = 30000,
                        control = list(adapt_delta = .99, max_treedepth = 12))

posterior_test <- extract(test_model)

alphas <- colMeans(posterior_test[["z"]])

plotting <- tibble(Lap = 1:63,
                   alphas = alphas,
                   observations = hamilton_laps)

ggplot(plotting, mapping = aes(x = Lap)) + 
  geom_point(mapping = aes(y =observations, col = "Observations"))+
  geom_point(mapping = aes(y = alphas, col = "Latent States"))

```
## Get Data Ready

```{r}

hamilton_df <- austria_laps %>%
    filter(Driver == "HAM", is.na(PitOutTime) & is.na(PitInTime), LapTime < 100) %>%
    mutate(Compound_Code = as.integer(factor(Compound,levels = c("HARD","MEDIUM","SOFT"))))

hamilton_laps <- hamilton_df$LapTime

hamilton_compound <- hamilton_df$Compound_Code

hamilton_pit <- rep(0,length(hamilton_compound))

for(i in 2:length(hamilton_df$Stint)) {
    if(hamilton_df$Stint[i] != hamilton_df$Stint[i-1]) {
        hamilton_pit[i-1] <- 1
    }    
}                  

fuel.kg <- seq(110,1,length.out = length(hamilton_laps))

hamilton_map <- sort(unique(hamilton_compound))
hamilton_used <- length(hamilton_map)

# Data for stan code
hamilton_data_base <- list(TT = length(hamilton_laps), y = hamilton_laps, C = 3, Compound = hamilton_compound, C_used = hamilton_used, compound_map = hamilton_map, Pit = hamilton_pit, z_reset0 = 69, v_reset0 = 0, sdo0 = .3, fuel_mass = fuel.kg)

hamilton_data_Ext <- list(TT = length(hamilton_laps), y = hamilton_laps, C = 3, Compound = hamilton_compound, C_used = hamilton_used, compound_map = hamilton_map, Pit = hamilton_pit, z_reset0 = c(69.5,69,68.5), v_reset0 = 0, sdo0 = .3, fuel_mass = fuel.kg)

```



```{r}

stan_results_base <- stint_CV("full_race_1driver_and_fuel_base.stan", hamilton_data_base)

stan_results_Ext1 <-stint_CV("full_race_1driver_and_fuel_Extension1.stan", hamilton_data_Ext)

stan_results_Ext2 <- stint_CV("full_race_1driver_and_fuel_Extension2.stan", hamilton_data_Ext)

arima_results <- arima_cv(hamilton_laps, hamilton_pit)

```


```{r}

stan_results_base[[1]]

stan_results_Ext1[[1]]

stan_results_Ext2[[1]]

arima_results[[1]]

results_tbl <- tibble(Arima = arima_results[[1]],
                      Base_Model = stan_results_base[[1]],
                      Extension_1 = stan_results_Ext1[[1]],
                      Extension_2 = stan_results_Ext2[[1]])

write.csv(results_tbl, file = "Model_Selection_Results.csv")

```

