---
title: "Cross Validation Study for a Single Stint"
author: "Cole Cappello"
date: "9-7-25"
output: pdf_document
---

```{r, include=FALSE}

set.seed(434)

library(mosaic)
library(tidyverse)
library(effects)
library(ggResidpanel)
library(catstats2)
library(Matrix)
library(KFAS)
library(rstan)

options(mc.cores = parallel::detectCores())

```


# Simulation and Code

Here we'll simulate from the model

\begin{align}
	y_t &= \alpha_t + \epsilon_t \\
	\alpha_{t+1} &= \alpha_t + \nu_t + \omega_t \\ 
	\nu_{t+1} &= \nu_t + \beta
\end{align}

where \( \epsilon_t \sim N(0,.3) \), \( \beta = .01 \), $\nu_1 = .01$, $\omega_t \sim N(0,.1)$, and $\alpha_1 = 96$.

```{r,results='hide'}


# Number of states
N <- 25

# Beta
beta <- .01

# Initialize nu vector
nu <- rep(NA, N)
nu[1] <- .01

# Initialize alpha vector
alpha <- rep(NA, N)
alpha[1] <- 96

# Initialize process errors
omega <- rnorm(n = N, mean = 0, sd = .1)

# Store true latent states
for(i in 2:(length(alpha))) {    
    alpha[i] <- alpha[i-1] + nu[i-1] + omega[i]
    nu[i] <- nu[i-1] + beta
}    

# Vector of observation errors
epsilon <- rnorm(n = N, mean = 0, sd = .3)

# Simulated observations
y <- alpha + epsilon

```


# Forward Chaining Cross-Validation

The problem with Cross-Validation in this setting is that we have a pretty small sample size.  Typically, a stint will be between 10 and 25 laps. For the 25 lap case above, the way I'd like to implement CV will be like so:

fold 1: training [laps 1-20], test [lap 21]

fold 2: training [laps 1-21], test [lap 22]

fold 3: training [laps 1-22], test [lap 23]

fold 4: training [laps 1-23], test[lap 24]

fold 5: training [laps 1-24], test [lap 25]

This seems like the best method out of those that I have come across for doing cross validation on a small sample size time series.

I could also potentially decrease the initial training size (maybe 15 instead of 20).

To estimate the one step ahead prediction, I have included code in the generated quantities block that will draw a predicted state and predicted observation for each posterior draw.

Lastly, the summary statistic we'll use is the Squared Prediction Error:
\[
(y_{T+1} - \hat{y}_{T+1})^2
\]

where $\hat{y}_{T+1}$ is the mean of the y_pred generated quantities.


## CV Function Definition

```{r,results='hide',warning=FALSE}

## This chunk will define a function that takes in a stan model and then performs the above cross validation schem.  This will hopefully make it easier to do model selection.

stint_CV <- function(model_string, data_list) {

    # Number of folds
    K <- 5

    stan.models <- vector("list", length = K)
    posteriors <- vector("list", length = K)

    # Vector to Store one-step ahead predictions
    ypred_hat <- rep(NA, K)

    for(i in 1:K) {

        # Update data vector to allow for next step prediction
        data_list[["TT"]] <- length(y[1:(i+N-K-1)])
        data_list[["y"]] <- y[1:(i+N-K-1)]
        
        # Fit model on subset of data then get one step ahead prediction
        stan.models[[i]] <- stan(file = model_string,
                                 data = data_list,
                                 chains = 3, iter = 30000,
                                 control = list(adapt_delta = .99))
        posteriors[[i]] <- extract(stan.models[[i]])
        ypred_hat[i]  <- mean(posteriors[[i]]$y_pred)
        
    }    

    # Calculate Summary Statistic
    MSPE <- mean((y[(N-K+1):N] - ypred_hat)^2)
    RMSPE <- sqrt(MSPE)

    plot_cv <- plot(x = ypred_hat, y = y[(N-K+1):N],
                    xlab = "Predictions",
                    ylab = "True Observations")

    return(list(RMSPE, plot_cv))

}

# list(TT = length(y[1:(i+N-K-1)]), y = y[1:(i+N-K-1)], z0 = 96, v0 = .01, sdo0 = .3)

```


## Cross Validation for Various Models 

Here are the 3 models that we'll do cross validation for:

\begin{align}
	y_t &= \alpha_t + \epsilon_t \\
	\alpha_{t+1} &= \alpha_t + \nu
\end{align}

\begin{align}
	y_t &= \alpha_t + \epsilon_t \\
	\alpha_{t+1} &= \alpha_t + \nu + \omega_t 
\end{align}

\begin{align}
	y_t &= \alpha_t + \epsilon_t \\
	\alpha_{t+1} &= \alpha_t + \nu_t + \omega_t \\ 
	\nu_{t+1} &= \nu_t + \beta
\end{align}

```{r,results='hide',warning=FALSE}

# Model 1 CV
data_stan1 <- list(TT = length(y), y = y, z0 = 96)

cv1 <- stint_CV("Linear_SSM.stan",data_stan1)

# Model 2 CV
data_stan2 <- list(TT = length(y), y = y, z0 = 96)

cv2 <- stint_CV("Basic_SSM2.stan", data_stan2)

# Model 3 CV
data_stan3 <- list(TT = length(y), y = y, z0 = 96, v0 = .01, sdo0 = .3)

cv3 <- stint_CV("Increasing_Slope_SSM.stan", data_stan3)

```

```{r}
# Model 1 RMSPE
cv1[[1]]

# Model 2 RMSPE
cv2[[1]]

# Model 3 RMSPE
cv3[[1]]

```


# Conclusion

The full model performs best here most of the time, which makes sense since that is the model that we are simulating from. Overall, I think the important thing here is that it gives a framework to do cross-validation for model selection on real data.


