---
title: "Initial Attempts at PMCMC"
author: "Cole Cappello"
date: "07=04-2025"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(message = FALSE)

set.seed(123)

knitr::opts_chunk$set(echo = TRUE)
library(mosaic)
library(tidyverse)
library(effects)
library(ggResidpanel)
library(catstats2)
library(Matrix)
library(KFAS)
library(rstan)

calc_rMSE <- function(alpha,estimate) {
  n <- length(alpha)
  rMSE <- 0
  for(i in 1:n){
    rMSE <- rMSE + (estimate[i] - alpha[i])^2
  }
  rMSE <- sqrt(rMSE/n)
  return(rMSE)
}

```

# Linearly Increasing Slope

What if we just let the slope increase linearly with each step to capture some of that increasing decay rate behavior?

Here we'll simulate from the model

\begin{align}
	y_t &= \alpha_t + \epsilon_t \\
	\alpha_{t+1} &= \alpha_t + \nu_t + \omega_t \\ 
	\nu_{t+1} &= \nu_t + \beta
\end{align}

where \( \epsilon_t \sim N(0,.3) \), \( \beta = .01 \), $\nu_1 = .01$, $\omega_t \sim N(0,.1)$, and $\alpha_1 = 96$.

```{r}


# Number of states
N <- 25

# Beta
beta <- .01

# Initialize nu vector
nu <- rep(NA, N)
nu[1] <- .01

# Initialize alpha vector
alpha <- rep(NA, N)
alpha[1] <- 96

# Initialize process errors
omega <- rnorm(n = N, mean = 0, sd = .1)

# Store true latent states
for(i in 2:(length(alpha))) {    
    alpha[i] <- alpha[i-1] + nu[i-1] + omega[i]
    nu[i] <- nu[i-1] + beta
}    

# Vector of observation errors
epsilon <- rnorm(n = N, mean = 0, sd = .3)

# Simulated observations
y <- alpha + epsilon

```

The Increasing_Slope_SSM.stan file fits the above model.  Increasing_Slope_SSM_Parameterized.stan fits a parameterized version of the model to try to help with divergences.

```{r}

data_stan <- list(TT = length(y), y = y, z0 = 96, v0 = .01, sdo0 = .3)

stan_test <- stan(file = "Increasing_Slope_SSM_Parameterized.stan",
                  data = data_stan,
                  chains = 3, iter = 30000,
                  control = list(adapt_delta = .99))

states <- colMeans(extract(stan_test, pars = c("z"))[[1]])

plot(x = states, y = alpha, 
     xlab = "Estimated States",
     ylab = "True States")
abline(0,1)

# Include a plot of time vs. estimated state, true state, and observations
time <- seq(from = 1, to = N, by = 1)

z_stan <- extract(stan_test, pars = c("z"))[[1]]
z_CIl <- apply(z_stan, 2, quantile, probs=0.025)
z_CIu <- apply(z_stan, 2, quantile, probs=0.975)

df1 <- tibble(time = time,
              alpha = alpha,
              alphahat = states,
              observations = y,
              z_CIl = z_CIl,
              z_CIu = z_CIu)


df_long <- df1 %>%
  pivot_longer(cols = c(alpha, alphahat,observations), names_to = "type", values_to = "value")

ggplot(df_long, aes(x = time, y = value, color = type)) +
    geom_point() +
    geom_ribbon(aes(ymin = z_CIl, ymax = z_CIu), fill = "black",color = "gray", alpha = .1)


# Estimated Observation Error:
mean(extract(stan_test, pars = c("sdo"))[[1]])

# Estimated Process Error
mean(extract(stan_test, pars = c("sdp"))[[1]])

calc_rMSE(alpha,states)
calc_rMSE(alpha, y)

```



# PMCMC

```{r}

library(nimble)
library(nimbleSMC)

code <- nimbleCode({
    # priors
    sdo ~ dnorm(sdo0, .1)
    sdp ~ dnorm(sdp0, .1)
    b ~ dnorm(0,.1)

    # Model
    v[1] ~ dnorm(v0,.1)

    for(t in 2:TT){
        v[t] <- v[t-1] + b
    }   
    
    z[1] ~ dnorm(z0,.1)

    for(t in 2:TT){
        z[t] ~ dnorm(z[t-1] + v[t-1],sdp)
    }    

    for(t in 1:TT){
        y[t] ~ dnorm(z[t],sdo)
    }    
})    

constants <- list(TT = 25,
                  sdo0 = .3,
                  sdp0 = .1,
                  v0 = 0,
                  z0 = 96)

data <- list(y = y)

inits <- list(sdo = .3,
              sdp = .1,
              b = 0,
              v = rep(0,N),
              z = rep(96,N))

deg.model <- nimbleModel(code = code,constants = constants, data = data,inits = inits)

compileNimble(deg.model)

deg.conf <- configureMCMC(deg.model, nodes = NULL,print = TRUE)

deg.conf$addSampler(target = c('sdo','sdp','b',"v[1]"),
                    type = "RW_PF_block",
                    control = list(latents = 'z',
                                   pfType = "bootstrap",
                                   pfNparticles = 2000))

deg.MCMC <- buildMCMC(deg.conf)

deg.comp <- compileNimble(deg.MCMC)

#deg.out <- nimbleMCMC(model = deg.comp, nchains = 2,
#                      summary = TRUE, monitors = TRUE

samples <- runMCMC(deg.comp, niter = 10000, nburnin = 1000, nchains = 2,summary = TRUE)

# Get the full summary table
summary_df <- as.data.frame(samples$summary$all.chains)

# Pull out just the z states by name
z_summary <- summary_df[grep("^z\\[", rownames(summary_df)), ]

alphahat <- z_summary[,1]

df <- tibble(time = seq(1,N,1),
             alphahat = alphahat,
             alpha = alpha,
             observations = y)

df_long1 <- df %>%
  pivot_longer(cols = c(alpha, alphahat,observations), names_to = "type", values_to = "value")

ggplot(df_long1, aes(x = time, y = value, color = type)) +
    geom_point() 
#    geom_ribbon(aes(ymin = z_CIl, ymax = z_CIu), fill = "black",color = "gray", alpha = .1)

calc_rMSE(alpha,alphahat)


```




```{r}

library(nimble)
library(nimbleSMC)

code <- nimbleCode({
    # priors
    sdo ~ dnorm(.3, 10)
    sdp ~ dnorm(.1, 10)

    b ~ dnorm(0,.1)


    z[1] ~ dnorm(z0,.1)

    for(t in 2:TT){
        z[t] ~ dnorm(z[t-1] + b,sdp)
    }    

    for(t in 1:TT){
        y[t] ~ dnorm(z[t],sdo)
    }    
})    

constants <- list(TT = 25,
#                  sdo0 = .3,
#                  sdp0 = .1,
#                  v0 = 0,
                  z0 = 96)

data <- list(y = y)

inits <- list(sdo = .3,
              sdp = .1,
              b = 0,
              z = rep(96,N))

deg.model <- nimbleModel(code = code,constants = constants, data = data,inits = inits)

compileNimble(deg.model)

deg.conf <- configureMCMC(deg.model,print = TRUE)

deg.conf$addSampler(target = c('sdo','sdp','b'),
                    type = "RW_PF_block",
                    control = list(latents = 'z',
                                   pfType = "bootstrap",
                                   pfNparticles = 1000))

deg.MCMC <- buildMCMC(deg.conf)

deg.comp <- compileNimble(deg.MCMC)

#deg.out <- nimbleMCMC(model = deg.comp, nchains = 2,
#                      summary = TRUE, monitors = TRUE

samples <- runMCMC(deg.comp, niter = 10000, nburnin = 1000, nchains = 2,summary = TRUE)

# Get the full summary table
summary_df <- as.data.frame(samples$summary$all.chains)

# Pull out just the z states by name
z_summary <- summary_df[grep("^z\\[", rownames(summary_df)), ]

alphahat <- z_summary[,1]

df <- tibble(time = seq(1,N,1),
             alphahat = alphahat,
             alpha = alpha,
             observations = y)

df_long1 <- df %>%
  pivot_longer(cols = c(alpha, alphahat,observations), names_to = "type", values_to = "value")

ggplot(df_long1, aes(x = time, y = value, color = type)) +
    geom_point() 
#    geom_ribbon(aes(ymin = z_CIl, ymax = z_CIu), fill = "black",color = "gray", alpha = .1)

calc_rMSE(alpha,alphahat)


```

# SMC Run treating parameters as fixed

```{r}



#SMCfunction <- function(parSdo, parSdp, parB) {

parSdo <- .3
parSdp <- .1
parB <- .01

    code <- nimbleCode({

        z[1] ~ dnorm(z0,5)

        for(t in 2:TT){
            z[t] ~ dnorm(z[t-1] + b,sdp)
        }    

        for(t in 1:TT){
            y[t] ~ dnorm(z[t],sdo)
        }    
    })    

    constants <- list(TT = 25,
                      sdo = parSdo,
                      sdp = parSdp,
                      b = parB,
                      z0 = 96)

    data <- list(y = y)

    inits <- list(z = rep(96,N))

    SMC.model <- nimbleModel(code = code,constants = constants, data = data,inits = inits)

    cSMC.model <- compileNimble(SMC.model)

    bootFilter <- buildBootstrapFilter(model = SMC.model, nodes = "z", control = list(saveAll = TRUE))

    cBootFilter <- compileNimble(bootFilter, project = cSMC.model)

    cBootFilter$run(m = 1000)

    head(as.matrix(cBootFilter$mvEWSamples))

#}    

    
    #----------------------------------------------------------------------#
    
    deg.conf <- configureMCMC(deg.model,print = TRUE)

    deg.conf$addSampler(target = c('sdo','sdp','b'),
                        type = "RW_PF_block",
                        control = list(latents = 'z',
                                       pfType = "bootstrap",
                                       pfNparticles = 1000))

    deg.MCMC <- buildMCMC(deg.conf)

    deg.comp <- compileNimble(deg.MCMC)

                                        #deg.out <- nimbleMCMC(model = deg.comp, nchains = 2,
                                        #                      summary = TRUE, monitors = TRUE

    samples <- runMCMC(deg.comp, niter = 10000, nburnin = 1000, nchains = 2,summary = TRUE)

                                        # Get the full summary table
    summary_df <- as.data.frame(samples$summary$all.chains)


```




# Attempts at PMCMC for the Full Model


## Simulation Code and Full Race Model with Fuel Loss

We'll create a fuel.kg vector which starts at 110 (this is the max amount of fuel allowed by regulation) and decreases linearly to 0 at the end of the race. We are going to use a similar model to the above but add a covariate to the observation equation for fuel. In order to estimate gamma correctly, it will be critical to set a proper prior for the initial latent $\alpha_1 = \alpha.reset[compound_{t=1}]$. Ultimately, it needs to be roughly 1-1.5 seconds less than the initial observation $y_1$, but in the stan code we'll set the prior to be centered on the true value of 95.

\begin{align}
	y_t &= \alpha_t + \gamma*fuel.kg_t +\epsilon_t \\
	\alpha_{t+1} &= 
	\begin{cases}
		\alpha_t + \nu_t + \omega_t & pit_t = 0 \\ 
		\alpha.reset[compound_t] & pit = 1
	\end{cases} \\	
	\nu_{t+1} &= 
	\begin{cases}
		\nu_t + \beta[compound_t] & pit_t = 0 \\
		\nu.reset & pit = 1
	\end{cases}	
\end{align}

where \( \epsilon_t \sim N(0,.2) \), \( \beta[Hard] = .002 \), $\beta[Medium] = .01$, $\beta[Soft] = .02$, $\nu_1 = \nu.reset = 0$, $\omega_t \sim N(0,.15)$, and $\alpha_1 = \alpha.reset[compound_{t=1}]$.  We'll let $\alpha.reset$ be 95, 94.5, and 94 for hard, medium, and soft tires resepectively. In this simulation, we will start on hard tires, pit on lap 30 for soft tires, then pit on lap 40 for medium tires. 




```{r}

# Number of states/laps
N <- 60

# Tire Compound index 1 - Hard, 2 - Medium, 3 - Soft
compound <- c(rep(1,times = 30), rep(3, times = 10),rep(2, times = 20))

# Fuel Mass in kg
fuel.kg <- seq(from = 110, to = 0, length.out = N)

# Gamma 
gamma <- .01


# Pit Index indicates with a 1 if a new set of tires is put on the next lap
pit <- rep(0, N)
pit[30] <- 1
pit[40] <- 1

# Beta is now a vector with the index representing each compound
beta <- rep(0,3)
beta[1] <- .002
beta[2] <- .01
beta[3] <- .02

# Reset vector stores the true pace of a new tire for each compound
reset.alpha <- rep(0,3)
reset.alpha[1] <- 95
reset.alpha[2] <- 94.5
reset.alpha[3] <- 94

# Reset value for nu
reset.nu <- 0
    
# Initialize nu vector
nu <- rep(NA, N)
nu[1] <- .01

# Initialize alpha vector
alpha <- rep(NA, N)
alpha[1] <- reset.alpha[1]

# Initialize process errors
omega <- rnorm(n = N, mean = 0, sd = .1)

# Store true latent states
for(i in 2:(length(alpha))) {
    if(pit[i-1] == 1) {
        alpha[i] <- reset.alpha[compound[i]] + omega[i]
        nu[i] <- reset.nu
    } else {
    alpha[i] <- alpha[i-1] + nu[i-1] + omega[i]
    nu[i] <- nu[i-1] + beta[compound[i-1]]
    }
}    

# Vector of observation errors
epsilon <- rnorm(n = N, mean = 0, sd = .2)

# Simulated observations
y <- alpha +  epsilon #+ fuel.kg*gamma 

```



## Attempt to Wrtie the Full Model in Nimble


```{r}


library(nimble)
library(nimbleSMC)

code <- nimbleCode({
    
    # priors
    sdo ~ dnorm(sdo0, sd = .1)
    sdp ~ dnorm(sdp0, sd = .1)
    b[1] ~ dnorm(.002, sd = .1)
    b[2] ~ dnorm(.01, sd = .1)
    b[3] ~ dnorm(.02, sd = .1)
    v_reset ~ dnorm(0,sd = .1)
    z_reset[1] ~ dnorm(95,sd = .1)
    z_reset[2] ~ dnorm(94.5,sd = .1)
    z_reset[3] ~ dnorm(94,sd = .1)
    
    # Model
    v[1] <- v_reset

    for(t in 2:TT){
        
        v[t] <- (1 - pit[t]) * (v[t-1] + b[compound[t]]) + pit[t] * (v_reset)

    }   
    
    z[1] ~ dnorm(z0,.1)

    for(t in 2:TT){
        
        z[t] ~ dnorm((1 - pit[t-1]) * (z[t-1] + v[t-1]) + pit[t-1] * z_reset[compound[t]], sd = sdp)
    }
    
    for(t in 1:TT){
        y[t] ~ dnorm(z[t],sd = sdo)
    }    
})    

constants <- list(TT = 60,
                  sdo0 = .3,
                  sdp0 = .1,
                  v0 = 0,
                  z0 = 96,
                  pit = pit,
                  compound = compound)

data <- list(y = y)

inits <- list(sdo = .3,
              sdp = .1,
              b = rep(0,3),
              v = rep(0,N),
              z = rep(96,N))

deg.model <- nimbleModel(code = code,constants = constants, data = data,inits = inits)

compileNimble(deg.model)

deg.conf <- configureMCMC(deg.model, nodes = NULL,print = TRUE)

deg.conf$addSampler(target = c('sdo','sdp','b',"v_reset","z_reset"),
                    type = "RW_PF_block",
                    control = list(latents = 'z',
                                   pfType = "bootstrap",
                                   pfNparticles = 2000))

deg.MCMC <- buildMCMC(deg.conf)

deg.comp <- compileNimble(deg.MCMC)

#deg.out <- nimbleMCMC(model = deg.comp, nchains = 2,
#                      summary = TRUE, monitors = TRUE

samples <- runMCMC(deg.comp, niter = 10000, nburnin = 1000, nchains = 2,summary = TRUE)

# Get the full summary table
summary_df <- as.data.frame(samples$summary$all.chains)




```



